Quantifying passive kelp rafting phenomena and connectivity of nearshore communities

### 2020-01-23 thesis_chapter_1.Rmd
updated rmd file for chapter 1. added to coefficients section

The need to understand the effects of oceanographic conditions affecting trajectory of floating objects has largely been borne from the maritime industry. The applications in the maritime industry include locating lost cargo, locating naval and plane deris, search and rescue, and the hydrodynamic effects on naval architecture.

The trajectory of passivley drifting objects on the sea surface is influenced by multiple factors, such as water currents, atmospheric wind, wave motion, wave induced currents, gravitaional force and buoyancy force. To complicate matters, the previously mentioned factors do not act independently of one another but instead influence one another. Futhermore, the gravitational and buoyancy forces on the object are determined by the objects shape. Therefore, all these factors need to be taken into account when modelling trajectory of drifting objects. Given the local wind, surface current, and the shape and buoyancy of the object is known, it is possible to estimate trajectory by the equation $$V_{drift}=V_{curr} + V_{rel}$$,

where $V_{curr}$ represents the current velocity relative to the earth, and $V_{rel}$ represents the object drift velocity relative to the ambient water. Ocean currents are determined by two components: the surface current (including the effects of Ekman drift, baroclinic motion, tidal and inertial currents) and Stokes drift induced by waves. The assumption made is $V_{curr}$ influences all floating objects in the same manner, regardless of shape or size. Therefore, $V_{curr}$ is equated with the surface current obtained from a numerical ocean model which has been paraterised by wind velocity. The effects of $V_{rel}$ on a floating object is driven by wind and wave forces which is dependent on the shape and size of the floating object.

## Wind effects on drifting objects in the oceans

The effects that wind produce on floating objects is sometimes referred to as "windage factors", and is difficult to accurately and empirically describe. The difficulty is due to accurate measurements needed for the current velocity, wind velocity, wave height and wave direction. Furthermore, these factors do not act independently and instead influence one another, which compounds the complexity of calculating wind effects on drifting objects. To overcome this challenge, measurements of current and wind velocity, and wave height and direction are measured along the objects trajectory for a wide range of conditions.

@@

$V_l$ is the velocity of the object relative to the water due to effects of wind force directly on the object and can expressed by $$V_l = w U_10$$
where $w$ is a linear 'windage' or 'leeway' coefficient and $U_10$ is the wind speed at 10m height. The study by @griffin2017 used a 10m height to model drifing airplane debris, however this may be adjusted according to drifting object being modelled. The nature of $w$ (linear windage) is dependent on the object, for instance yachts tend to drift in a particular orientation to the wind, and so the angle of the drifting object is dependent on said orientation. Furthermore, drifting object which are more wind exposed and are less submerged will have high windage. Due to these factors, and to allow for a non-zero angle between $V_l$ and $U_10$, $w$ (linear windage) can be a complex number.

$V_W$ is the velocity of the object as a result of direct wave force, and is dependent on the size and shape of the object. In instances when wave-wave interactions result in an small, unbalanced force then $V_W$ is assumed to be zero. Also, objects may be able to absorb wave energy which can result in a non-zero $V_W$ value. Since surface-waves are wind driven, the $V_W$ is calculated by $$V_W = f(U_10)$$
# Particle dispersion modelling
## Particle dispersion models

$$u'= u + \frac{\delta tu_t}{2}$$ $$\upsilon'=\upsilon + \frac{\delta t\upsilon_t}{2}$$
and $u$, $\upsilon$ are orthogonal velocity components, $t$ is time, $\delta t$ is the model time step, $u_\chi$, $u_y$, $\upsilon_y$ are the $u$ and $\upsilon$ spatial velocity gradients, and $u_t$, $\upsilon_t$ are temporal gradients. Horizontal diffusion was modeled as a random walk with separate longitudinal and lateral coefficients set to simulate random turbulence. The distance increments moved by the particle at each time step is calculated as:
$$\Delta X = R_N \sqrt{6E_1\delta t}$$ $$\Delta y = R_N\sqrt{6E_2\delta t}$$
where $R_N$ is a random number in the uniform range (-1,1) and $E_1,E_2$ are the longitudinal and lateral eddy diffusivities, respectively.

Dispersal is recognised as important driver of biodiversity, composition and structure of ecological systems in the marine environment [@bernardes2018;@helmuth1994;@highsmith1985;@macarthur2001;@jackson2010]. These processes are largely dependent on currents for dispersal of migrant populations which ultimately promote connectivity of marine ecosystems [@bernardes2018;@mackas1985;@mackas1985;@zakas2009]. This is particularly true for organisms which lack pelagic larvae, and are reliant on other modes of transport to reach new habitats through passive modes of dispersal from anthropogenic and natural sources. Anthropogenic sources include marine litter and structural debris while examples of organic sources include pumice and macroalgae in the form of kelp-rafts. Kelp-rafts have been identified as an important mode of passive dispersal of marine organisms, such as invertebrates [@helmuth1994; whichmann2012], epiphytes [@], grazers [@nikula2010], invasive species [@lewis2005] and macroalgae themselves [@macaya2005] which either make landfall or are transported offshore where they eventually sink. A kelp-raft consists of an entanglement of macroalgae (one or multiple species) which have been dislodged from the benthic environment through hydrodynamic forces (mostly storms) and are positively buoyant by means of gas-filled pneumatocysts and/or reproductive organs. Kelp-rafts are capable of travelling vast distances [@] and are considered important dispersal vectors in temperate latitudes, such as the Southern California Bight [@hobday2000] and the Northern Baltic Sea [@].

# Buoyancy of drifting objects in the oceans

@@

Once the fractions of submerged and non-submerged portions of the object have been calculated the next calculation relevant to this study would be to calculate the weight needed to cause the object to sink or as least be completely submerged. This will relate to epiphyte loading on kelp rafts in order to determine raft longevity. If a weight is placed on the object in the conceptual model that has enough weight to fully submerge (but not sink) the object then an addition buoyancy force is needed, which will be represented by $BF_2$ which is equal to the additional mass that has been added which can be expressed as $mg$. Based on the previous calculations and the laws of physics we know that: $$W_{object} = mg = BF_2 = W_{add-liq-displaced}$$ The weight of the additional liquid displaced can now be calculated by substituting the density and volume of the additional displaced liquid along with the gravitational constant $g$ which is expressed as: $$W_{AL} = \rho_{l}V_{AL}g$$
where the subscript $AL$ represents the components of *additional liquid displaced*. The $V$ can also represent the change in volume ($\Delta V$) and so the expression can be expanded to: $$W_{AL} = \rho(\Delta V) g$$} where $\Delta V$ represents the change in volume of the object, i.e $y$. To estimate the amount of input into the model the authors used a scaled approach to define the release of particles, and where based on previous research performed by @halpern2008.

>>>>

The need to understand the effects of oceanographic conditions affecting trajectory of floating objects has largely been borne from the maritime industry. The applications in the maritime industry include locating lost cargo, locating naval and plane debris, search and rescue, and the hydrodynamic effects on naval architecture.

The trajectory of passively drifting objects on the sea surface is influenced by multiple factors, such as water currents, atmospheric wind, wave motion, wave induced currents, gravitational force and buoyancy force. To complicate matters, the previously mentioned factors do not act independently of one another but instead influence one another. Furthermore, the gravitational and buoyancy forces on the object are determined by the objects shape. Therefore, all these factors need to be taken into account when modelling trajectory of drifting objects. Given the local wind, surface current, and the shape and buoyancy of the object is known, it is possible to estimate trajectory by the equation $$V_{drift}=V_{curr} + V_{rel}$$

where $V_{curr}$ represents the current velocity relative to the earth, and $V_{rel}$ represents the object drift velocity relative to the ambient water. Ocean currents are determined by two components: the surface current (including the effects of Ekman drift, baroclinic motion, tidal and inertial currents) and Stokes drift induced by waves. The assumption made is $V_{curr}$ influences all floating objects in the same manner, regardless of shape or size. Therefore, $V_{curr}$ is equated with the surface current obtained from a numerical ocean model which has been parameterised by wind velocity. The effects of $V_{rel}$ on a floating object is driven by wind and wave forces which is dependent on the shape and size of the floating object. Previous research has shown that the effects of waves on drifting objects are negligible when the length of the object is less than the wave length; and increase significantly when the lengths are approximately the same [@grue1993; @hodgins1998]. Furthermore, current velocity $V_{curr}$ affects on drifting objects is considered to be negligible in this approach and ideally should be taken into account. Studies investigating kelp rafts have identified surface current velocity (as a function of wind velocity) as an important influence in determine trajectory [@harrold1989; @macaya2005; @hobday2000a; @hobday2000b]. Therefore, the objects motion/trajectory relative to the wind must be taken into account.

## Windage factors

An objects motion relative to the wind is sometimes referred to as "windage factors" or "leeway drift, and is difficult to accurately and empirically describe. The difficulty is due to accurate measurements needed for the current velocity, wind velocity, wave height and wave direction. Furthermore, these factors do not act independently and instead influence one another, which compounds the complexity of calculating wind effects on drifting objects. To overcome this challenge, measurements of current and wind velocity, and wave height and direction are measured along the objects trajectory for a wide range of conditions.

@@

$V_l$ is the velocity of the object relative to the water due to effects of wind force directly on the object and can expressed by $$V_l = w U_10$$
where $w$ is a linear 'windage' or 'leeway' coefficient and $U_10$ is the wind speed at 10m height. The study by @griffin2017 used a 10m height to model drifting airplane debris, however this may be adjusted according to drifting object being modeled. The nature of $w$ (linear windage) is dependent on the object, for instance yachts tend to drift in a particular orientation to the wind, and so the angle of the drifting object is dependent on said orientation. Furthermore, drifting object which are more wind exposed and are less submerged will have high windage. Due to these factors, and to allow for a non-zero angle between $V_l$ and $U_10$, $w$ (linear windage) can be a complex number.

$V_W$ is the velocity of the object as a result of direct wave force, and is dependent on the size and shape of the object. In instances when wave-wave interactions result in an small, unbalanced force then $V_W$ is assumed to be zero. Also, objects may be able to absorb wave energy which can result in a non-zero $V_W$ value. Since surface-waves are wind driven, the $V_W$ is calculated by $$V_W = f(U_10)$$

Important to note is that the effects of wind and waves on the trajectory of a drifting object may be influenced by the object's orientation. This has been noted in modelling drift trajectories of ships [@allen1999review; @hackett2006], and should be incorporated into the leeway drift component. The direct ($V_l$) and indirect ($V_s$ and $V_w$) effects do not act independently and instead influence one another. with regards to this, @griffin2017 recommend incorporating an effective windage factor which would represent the combined effects of Stokes Drift, leeway drift and wave forces. Although a common approach, this may not be necessary in all circumstances particularly if individual effects of factors. As mentioned previously, the magnitude of effects of waves and wind on a drifting object is dependent on its shape, size and buoyancy.

## Buoyancy of drifting objects in the oceans

@@

Once the fractions of submerged and non-submerged portions of the object have been calculated the next calculation relevant to this study would be to calculate the weight needed to cause the object to sink or as least be completely submerged. This will relate to epiphyte loading on kelp rafts in order to determine raft longevity. If a weight is placed on the object in the conceptual model that has enough weight to fully submerge (but not sink) the object then an addition buoyancy force is needed, which will be represented by $BF_2$ which is equal to the additional mass that has been added which can be expressed as $mg$. Based on the previous calculations and the laws of physics we know that: $$W_{object} = mg = BF_2 = W_{add-liq-displaced}$$ The weight of the additional liquid displaced can now be calculated by substituting the density and volume of the additional displaced liquid along with the gravitational constant $g$ which is expressed as: $$W_{AL} = \rho_{l}V_{AL}g$$
where the subscript $AL$ represents the components of *additional liquid displaced*. The $V$ can also represent the change in volume ($\Delta V$) and so the expression can be expanded to: $$W_{AL} = \rho(\Delta V) g$$} where $\Delta V$ represents the change in volume of the object, i.e $y$.

## Buoyancy factors

The buoyancy of objects at sea has been shown to be influeced by epibiont load [@hobday2000b; @rotha2011; @graiff2016; @macaya2016], which reduces drift times by reducing buoyancy with increasing biomass, which ultimatley leads to the object sinking. The bouyancy of the drifting object is determined by the growth rate of epiphytic species while drifting, which in turn will be depndent on environmental factors such as temperature and light which will vary along the objects trajectory Therefore, buoyancy should be parameterised by temporal epibiont biomass load.

# Particle dispersion modelling
## Particle dispersion models

$$u'= u + \frac{\delta tu_t}{2}$$ $$\upsilon'=\upsilon + \frac{\delta t\upsilon_t}{2}$$
and $u$, $\upsilon$ are orthogonal velocity components, $t$ is time, $\delta t$ is the model time step, $u_\chi$, $u_y$, $\upsilon_y$ are the $u$ and $\upsilon$ spatial velocity gradients, and $u_t$, $\upsilon_t$ are temporal gradients. Horizontal diffusion was modeled as a random walk with separate longitudinal and lateral coefficients set to simulate random turbulence. The distance increments moved by the particle at each time step is calculated as:
$$\Delta X = R_N \sqrt{6E_1\delta t}$$ $$\Delta y = R_N\sqrt{6E_2\delta t}$$
where $R_N$ is a random number in the uniform range (-1,1) and $E_1,E_2$ are the longitudinal and lateral eddy diffusivities, respectively. To estimate the amount of input into the model the authors used a scaled approach to define the release of particles, and where based on previous research performed by @halpern2008.

Dispersal is recognised as important driver of biodiversity, composition and structure of ecological systems in the marine environment [@bernardes2018;@helmuth1994;@highsmith1985;@macarthur2001;@jackson2010]. These processes are largely dependent on currents for dispersal of migrant populations which ultimately promote connectivity of marine ecosystems [@bernardes2018;@mackas1985;@mackas1985;@zakas2009]. This is particularly true for organisms which lack pelagic larvae, and are reliant on other modes of transport to reach new habitats through passive modes of dispersal from anthropogenic and natural sources. Anthropogenic sources include marine litter and structural debris while examples of organic sources include pumice and macroalgae in the form of kelp-rafts. Kelp-rafts have been identified as an important mode of passive dispersal of marine organisms, such as invertebrates [@helmuth1994; whichmann2012], epiphytes [@], grazers [@nikula2010], invasive species [@lewis2005] and macroalgae themselves [@macaya2005] which either make landfall or are transported offshore where they eventually sink. A kelp-raft consists of an entanglement of macroalgae (one or multiple species) which have been dislodged from the benthic environment through hydrodynamic forces (mostly storms) and are positively buoyant by means of gas-filled pneumatocysts and/or reproductive organs. Kelp-rafts are capable of travelling vast distances [@] and are considered important dispersal vectors in temperate latitudes, such as the Southern California Bight [@hobday2000a] and the Northern Baltic Sea [@].
